---
title: "Practical Machine Learning Project"
output: html_document
---

## Introduction 

This project looks to predict people's activities during exercise.
Training and testing data sets come from this source: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har).

## Initialization and loading required packages 
```{r initialization, results='hide', message=FALSE, warning=FALSE}
setwd("c:/Users/Owner/Desktop/PML-Project/")

require(caret)
require(randomForest)

setInternet2(TRUE)     # needed for downloading https urls

trainingUrl  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingUrl   <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainingDest <- "C:/Users/Owner/Desktop/PML-Project/data/pml.train.csv"
testDest     <- "C:/Users/Owner/Desktop/PML-Project/data/pml.test.csv"
```

## Download and Clean Data Sets
```{r downloadLoad}
if(!file.exists("data")){dir.create("data")}

download.file(trainingUrl, destfile=trainingDest)
download.file(testingUrl, destfile=testDest)

training <- read.csv(trainingDest,
                     na.strings=c("NA", "", "NULL"))
testing  <- read.csv(testDest,
                     na.strings=c("NA","", "NULL"))

dim(training); dim(testing)
```

Examination of the data reveals that many variables (columns) contain missing values. I remove these variables from both the training and test dataframes and also remove other unnecessary variables containing descriptive related information (i.e. index, names, etc ...). 

```{r cleanDatasets}
training <- training[, colSums(is.na(training)) == 0]
testing  <- testing[, colSums(is.na(testing)) == 0]

training <- training[,8:60]
testing  <- testing[,8:60]

dim(training); dim(testing)
```

To perform additional validation of our model, I create a validation dataset from the training data.

```{r createValidation}
inTrain    <- createDataPartition(training$classe,
                                  p=0.80,
                                  list=FALSE)
validation <- training[-inTrain,] 
training   <- training[inTrain,]
```

## Building a Model and evaluation

I'll try a random forest method to predict the activities classes.  
```{r buildModel}
set.seed(123)
modelFit <- randomForest(classe~., data=training)
modelFit
```
The model appeared to perform well (better than 99%) with an Out of Bag Error (OOB Error) of 0.38%.
I next cross validate the model using the validation data set.
```{r x-Validate}
prediction <- predict(modelFit, newdata=validation)
confMatrix <- confusionMatrix(prediction, validation$classe)
confMatrix
```
The model also produces low errors and is highly acurate (`r confMatrix$overall[[1]] * 100`%) against the validation set so I will use the the model to predict the activity classe for the test set.
```{r submission}
answers <- as.vector(predict(modelFit, newdata=testing))

pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}

pml_write_files(answers)
```
